A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-20 paper.tex
Tidy up section on Wiuf & Hein ARG
Rescale WH figure

\begin{figure}
\centering
% TODO update the source of this figure to make less tall. It's taking
% up far too much vspace at in native coords at the moment.
\includegraphics[width=0.75\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) showing the corresponding breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the samples, \noderef{a}, \noderef{b}, and \noderef{c}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{g} and \noderef{q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events have been retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure}

@@

%(see Appendix XXX) and has several features that we are unlikely to simulate or
%infer in practice. For example, by definition the samples contain no genetic information about
%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
%\textsf{q} as all regions of the genome have already reached a
%common ancestor separately (in nodes \textsf{k} and \textsf{p}).

% Third, there exist nodes such as
% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
% the breakpoint, although they retain the information that a breakpoint occurred between
% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

% reduce the annotation intervals to those that are ancestral to the
% sample (note that node $q$ is now ommitted because the MRCA of the sample
% has already been found for the full genome at node $p$).
% Tracking such
% ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
% is much more useful~\citep{mcgill2013graphml,arenas2013importance}
% and, indeed, all that is possible to
% have information about when inferring ARGs from contemporary samples.
% We refer to a gARG with inheritance annotations of this type as
% ``sample resolved''.

% Fig~\ref{fig-ancestry-resolution} is an instructive example
% from the literature \citep{wiuf1999recombination} and illustrates
% how a classical eARG (Fig~\ref{fig-ancestry-resolution}A) can be described as
% a gARG with edge annotations (Figs~\ref{fig-ancestry-resolution}B~and~C).

% There are two distinct annotation schemes possible, and the distinction is subtle
% and important. The first scheme is to use the annotations to describe the complete
% set of genomic intervals directly inherited by the descendant node from the ancestor
% node (Fig~\ref{fig-ancestry-resolution}B). The second is to use the edge annotations to
% describe the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
% of the sample that is propagated through
% the graph from descendant to ancestor (Fig~\ref{fig-ancestry-resolution}C).
% As we discuss in section XXX, this second scheme has many advantages
% in terms of efficiency.

%ARGs are inherently a retrospective structure defined in terms
%of a set of sample nodes (usually the leaves), and,
%as we discuss in section XXX, storing sample-resolved ancestry
%intervals (Fig~\ref{fig-ancestry-resolution}C) is generally much more
%useful than storing the parent-child intervals recording direct inheritance
%(Fig~\ref{fig-ancestry-resolution}B).
%Indeed, storing direct inheritance intervals for edges is really only of interest because of
%its correspondence with the classical approach of associating
%breakpoints with recombination nodes in an eARG (Fig~\ref{fig-ancestry-resolution}A).
%The eARG encoding is simply a more limited form, in which the only types of direct
%inheritance we can have are the entire genome, or the intervals
%at either side of a recombination breakpoint.

%Note that this example is a sample from the so-called \emph{``big'' ARG process}
%(see Appendix XXX) and has several features that we are unlikely to simulate or
%infer in practice. For example, by definition the samples contain no genetic information about
%nodes \textsf{g} and nodes \textsf{j} and so neither of those nodes are likely to be inferred.
%Likewise, is is unlikely we would have information to infer the ``grand MRCA''
%\textsf{q} as all regions of the genome have already reached a
%common ancestor separately (in nodes \textsf{k} and \textsf{p}).

% Third, there exist nodes such as
% \textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
% but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
% within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
% node \textsf{k}). In this case sample-resolved edge annotations lose the exact position of
% the breakpoint, although they retain the information that a breakpoint occurred between
% positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

% Given a gARG with direct-inheritance edge annotations
% (Fig~\ref{fig-ancestry-resolution}B),
%  we obtain the sample-resolved edge annotation
% (Fig~\ref{fig-ancestry-resolution}C) by traversing the graph backwards in time
% from sample nodes (\textsf{a}, \textsf{b}, \textsf{c}).
% Proceeding in topological order (such that an ancestor node is visited
% after all of its descendants) the sample-resolved material for a node is the
% union of the sample-resolved intervals on its inbound edges. We then annotate
% the outbound edges with the intersection of their direct-inheritance intervals
% and the just-computed sample-resolved intervals for the node.
% For example, node \textsf{i} in Fig~\ref{fig-ancestry-resolution}C
% carries sample-resolved material on the intervals $[0, 2)$ and $[4, 7)$. The edge
% joining \textsf{i} to its ancestor \textsf{k} then carries
% material on $[0, 2)$ since although \textsf{i} directly inherits interval
% $[0, 3)$ from \textsf{k} (Fig~\ref{fig-ancestry-resolution}B), only $[0, 2)$ is
% ancestral to the sample. For each node, we also keep track of the number of
% descendant samples over each genomic interval, and stop tracking that
% interval once the lineages from all samples have coalesced. For example,
% the edge from \textsf{k} to \textsf{n} omits the region $[0,2)$ because it has
% already fully coalesced (indicated by a fully black genomic region).

% In practice, the backwards-in-time Griffiths process that produces
% ARGs such as that shown in Figs~\ref{fig-ancestry-resolution}A~and~B
% is rarely simulated, because the resulting ``big ARG'' (Appendix XXX) is
% dominated by nodes such as \textsf{g} and \textsf{j} which contain information
% about inheritance in non-ancestral material, as well as nodes such as
% \textsf{o} and \textsf{q} which track details of ancestry even once the
% local genomic region has fully coalesced. Instead, simulating
% ARGs from the coalescent with recombination is usually done by
% running Hudson's algorithm to produce a sample-resolved ARG directly.
% Moreover, when inferring ARGs from genetic data, there is no information
% in the sample set that can be used to resolve these nodes, so that they
% are essentially unknowable.

% The process of converting a gARG with local-inheritance annotations
% to its sample-resolved equivalent is also known as ``simplification'',
% and is the key algorithm underlying recent
% advances in forwards-in-time
% simulation~\citep{kelleher2018efficient,haller2018tree}.
% It has many useful properties, as we discuss in the next section.

% It is worth noting the process of sample resolution can reveal that certain ARG
% nodes are \emph{redundant}: if they are removed and the edges adjusted to
% pas directly from the node's child to the node's parent, it does not reduce our
% knowledge of the samples' genetic ancestry. Removing nodes,
% both redundant ones and others, is part of a process we call
% ``simplification''.

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
``unary'' nodes in the local trees
(where a node has exactly one child over a particular genomic region).
% "unary regions" is jargon - just dropping it in here like this doesn't
% clarify. We can't assume that people are used to switching between
% graphs and trees.
% of ``unary'' regions for nodes
% (where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citep{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees.
}
\end{figure}

@@

% \noderef{a} in the leftmost tree (covering positions 0 to 30)
% Consider, for example, the path ancestral to node
% \noderef{a} in the leftmost tree (covering positions 0 to 30)
% in Fig.~\ref{fig-simplification}A.
% We can see from the graph that
% % Before \noderef{a} meets its common ancestor with the other samples at
% % node \noderef{q}, it passes through \noderef{e}, \noderef{j}, \noderef{m}, \noderef{n} and
% % \noderef{p}, which in this region of the genome are all unary nodes.
% % Node \noderef{e}

% Such unary nodes in a local tree are not usually considered important;
% % because they provide little information and are not
% % detectable by information from any given individual tree.
% nodes usually mark the points where samples coalesce in a
%  \emph{common} ancestor. For most purposes, we would therefore consider a path directly
% joining sample \noderef{a} to node \noderef{q} to be equivalent to the path
% containing the intermediate unary nodes.

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.
% something which is impossible to encode in an eARG.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question [CITE Ralph in prep.].

@@

% or nothing'': there is a gradation of detail
% about recombination, in terms of genomic location, timing,
% and the ancestral lineages involved.
% that vary levels of precision about reccomb

% The example in Fig.~\ref{fig-simplification} is usually
% referred to as ``fully simplified'', and is the default output
% of simulation programs like \msprime. While there is significant
% precision about recombination events lost

@@

% % Not sure where this goes, but there are some critical points
% % that we want to explain and not just throw out there in the
% % discussion.
% % A critical distinction between the eARG and gARG encodings
% % is the level of precision about the time and location of
% % recombination events that is required. In an eARG, we are
% % required to be absolutely precise about every recombination
% % [explain and contrast with gARG, referring to examples]

@@

% It is helpful to consider the difference between fully
% resolved binary trees and those that contain polytomies.

% The situation is in fact directly analogous: an ARG that
% contains fully resolved recombination information must have
% at most parents per node.

% At the very least, it is not reasonable for the \emph{data
% structure} to require complete precision in order to
% represent some less precise. This is a fundamental
% advantage of the gARG encoding over the classical eARG
% method: we are not forced to posit a precise cause (recombination
% event) for every effect (difference between adjacent trees).
% These differences in precision of recombination
% information are illustrated in the ARGs estimated by different
% methods, as we see in the next section.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{minichiello2006mapping,griffiths1996ancestral} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]

@@

% \section*{When do we need ``full'' ARGs?}
% % The full ancestral recombination graph (ARG) is a structure that encodes all
% % coalescence and recombination events resulting from the stochastic process of
% % the coalescent with recombination.
% \citet{brandt2021evaluation} define a ``full'' ARG as
% ``a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination'' (thus going against the general
% historial trend discussed in the XXX section).
% \citet{rasmussen2022espalier} have a less restrictive definition, in
% which a ``full'' ARG is one that contains complete information about
% recombination events, without necessarily being tied to a particular
% stochastic process.
% While [something positive], it is important to be clear on
% what the overall goal of such precise inferences are,
% what the limitations on when they can be made are,
% and their downstream utility in applications.

@@

% % Is it practical to estimate full ARGs?
% % [Not particularly \citep{shipilina2022origin} - but maybe ARGneedle
% % has done this for UKB? Need to check]

>>>>

\begin{figure}
\centering
% TODO update the source of this figure to make less tall. It's taking
% up far too much vspace at in native coords at the moment.
\includegraphics[width=0.6\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example graph
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) showing the corresponding breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the samples, \noderef{a}, \noderef{b}, and \noderef{c}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{g} and \noderef{q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events have been retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure}

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence
``unary'' nodes in the local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citep{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees.
}
\end{figure}

@@

@@

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
multiple---often indistinguishable---events into a single node.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.
% I don't think there's an actual manuscript or title to refer to here
% [CITE Ralph in prep.].

@@

@@

@@

% It is helpful to consider the difference between fully
% resolved binary trees and those that contain polytomies.
% The situation is in fact directly analogous: an ARG that
% contains fully resolved recombination information must have
% at two most parents per node.

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]

@@

@@
